---
title: "Morales-Castilla et al. (2015, TREE) Inferring biotic interactions from proxies."
author: "Vanessa, Gian-Marco, Thomas"
date: "June 16, 2015"
output: html_document
---

## Info

This is what ought to be the reproduction of the analyses presented in the article *Inferring biotic interactions from proxies* by Ignacio Morales-Castilla, Miguel G. Matias, Dominique Gravel, and Miguel B. Ara??jo (2015, TREE). The article can be found [here](http://www.sciencedirect.com/science/article/pii/S0169534715000774). Unfortuntely, reproducing thier analyses requires to go deep into two other studies, one about the Serengeti food web ([Baskerville et al. (2011, PLoS Computational Biology)](http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1002321)), and one about three Caribean coral reefs ([Roopnarine & Hertog (2013, Dataset Papers in Ecology)](http://www.hindawi.com/journals/dpis/2013/857470/abs/)).


## About



## Comments, errors & modifications

### Article text

* The number of potential links for the Cuban coral reef set should read as **53.130**, and *not* 53.361 because $n_{links}=n_{species}^2-n_{species}$. This number is based on the number of links found in the Cuban coral reef ($n_{guilds}=231$).
* It is irritating that the authors provide the total number of species found in all three coral reefs (265; Cayman, Cuba, Jamaica), instead of providing only the number of guilds found in the Cuban reef (231). The actual analysis is based on the **guilds** and *not* species. Therefore, the number of links in Fig. 1f is 53.130 (see paragraph above).

###  Data sets

* Some guild names have identical names, e.g., guild no. 235 and 236 are both named 'snapper'. I (Thomas) therefore changed the guild names to 'snapper a' and 'snapper b', respectively. I kept the original .csv file untuched and created a new one called '857470.item.2 - Trophic data for Cuba_sorted.csv'. We use this in our analyses. Here is a list of all guilds that have been modified as stated above:
    * 114, 117, 179 >> Angelfish a, Angelfish b, Angelfish c
    * 253, 255 >> Barracuda a, Barracuda b
    * 220, 257 >> Bigeye a, Bigeye b
    * 124, 125, 126, 158 >> Blenny a, Blenny b, Blenny c, Blenny d
    * 120, 122, 171, 175 >> Butterflyfish a, Butterflyfish b, Butterflyfish c, Butterflyfish d
    * 19, 20 >> Carnivorous micro-zooplankton a, Carnivorous micro-zooplankton b
    * 218, 223 >> Flounder a, Flounder b
    * 191, 206 >> Goatfish a, Goatfish b
    * 211, 212 , 246, 254, 260 >> Grouper a, Grouper b, Grouper c, Grouper d, Grouper e
    * 142, 144, 147, 192, 194, 195, 196, 199 >> Grunt a, Grunt b, Grunt c, Grunt d, Grunt e, Grunt f, Grunt g, Grunt h
    * 162, 222 >> Jacknife fish a, Jacknife fish b
    * 22, 23 >> Large omnivorous micro-zooplankton a, Large omnivorous micro-zooplankton b
    * 239, 248 >> Mackerel a, Mackerel b
    * 208, 250 >> Margate a, Margate b
    * 214, 216 >> Moray a, Moray b
    * 119, 127 >> Parrotfish a, Parrotfish b
    * 132, 133 >> Pufferfish a, Pufferfish b
    * 161, 204, 213, 228, 235, 236, 237, 238, 252 >> Snapper a, Snapper b, Snapper c, Snapper d, Snapper e, Snapper f, Snapper g, Snapper h, Snapper i
    * 140, 141, 143, 146 >> Squirrelfish a, Squirrelfish b, Squirrelfish c, Squirrelfish d
    * 128, 130 >> Trunkfish a, Trunkfish b
    * 148, 149, 164, 166 >> Wrasse a, Wrasse b, Wrasse c, Wrasse d
* In the .xlsx file '857470.item.2' (here used as '857470.item.2 - Trophic data for Cuba.csv') containing the trophic information of the Cuban coral reef some prey guild shave obscure prey names/values/IDs. For example, the prey of a baracuda is named '152153'. This does not make any sense. It seems likely that the authors forgot a comma and this baracuda feeds upon guilds number 152 and 153. I (Thomas) changed the .csv file for this. The file used in our analyses is called '857470.item.2 - Trophic data for Cuba_sorted.csv'. Here is a list of all guilds (number) whose prey items have been modified as stated above:
    * 253 (Barracuda a)  >> 152153 changed to 152, 153
    * 174 (Carnivorous fish I)  >> comma inserted after 100
    * 220 (Bigeye a)  >> comma inserted after 100
    * 115 (Atlantic spadefish) >> full stop replaced by comma between 20 and 22
    * 176 (Bass)  >> comma inserted after 100
    * 257 (Bigeye b)  >> comma inserted after 100
    * 258 (Blacktip shark)  >> comma inserted after 200
    * 158 (Blenny d)  >> comma inserted after 100
    * 125 (Blenny b)  >> comma inserted after 100
    * 171 (Butterflyfish c)  >> comma inserted after 100
    * 175 (Butterflyfish d)  >> comma inserted after 100
    * 122 (Butterflyfish b)  >> comma inserted after 100
    * 261 (Caribbean reef shark) >> comma inserted after 110, 132, 220, 250, 200
    * 224 (Coney grouper) >> comma inserted after 100
    * 123 (Damselfish)  >> comma inserted after 100
    * 173 (Flamefish)  >> comma inserted after 100
    * 223 (Flounder b)  >> comma inserted after 100
    * 191 (Goatfish a)  >> comma inserted after 100
    * 206 (Goatfish b)  >> comma inserted after 100
    * 206 (Goatfish b)  >> comma inserted after 100
    * 212 (Grouper b) >> 207196 replaced by 207, 196
    * 254 (Grouper d) >> comma inserted after 110
    * 260 (Grouper e) >> comma inserted after 100
    * 142 (Grunt a) >> comma inserted after 100
    * 144 (Grunt b) >> comma inserted after 100
    * 192 (Grunt d) >> comma inserted after 100
    * 196 (Grunt g) >> comma inserted after 100
    * 199 (Grunt h) >> comma inserted after 100
    * 221 (Hamlet) >> comma inserted after 100
    * 225 (Hamlets) >> comma inserted after 100
    * 169 (Hawkfish) >> comma inserted after 100
    * 232 (Jack) >> comma inserted after 100, 120; dublicates of 71, 72, 73, 74, 75, 79, 92, 93, 94, 95, 96, 97, 99, 98, 100 and 103 deleted
    * 155 (Hogfish) >> comma inserted after 100
    * 162 (Jacknife fish a) >> comma inserted after 100
    * 222 (Jacknife fish b) >> comma inserted after 100
    * 182 (Jawfish) >> 100103 changed to 100, 103
    * 239 (Mackerel a) >> comma inserted after 200
    * 248 (Mackerel b) >> comma inserted after 200
    * 250 (Margate b) >> comma inserted after 100
    * 214 (Moray a) >> 142213 changed to 142, 213
    * 150 (Nurse shark) >> comma inserted after 100
    * 205 (Palometa) >> comma inserted after 200
    * 165 (Porkfish) >> comma inserted after 100
    * 163 (Pufferfish b) >> comma inserted after 100
    * 217 (Red hind) >> comma inserted after 100
    * 219 (Scorpionfish) >> comma inserted after 100
    * 145 (Sharptail eel) >> comma inserted after 100
    * 198 (Sheepshead) >> comma inserted after 100
    * 152 (Slender Inshore Squid) >> comma inserted after 100  
    * 138 (Slippery dick) >> comma inserted after 100 
    * 235 (Snapper e) >> comma inserted after 100 
    * 236 (Snapper f) >> comma inserted after 100 
    * 252 (Snapper i) >> comma inserted after 100 
    * 168 (Soldierfish) >> comma inserted after 100 
    * 167 (Spotted drum) >> comma inserted after 100 
    * 197 (Spotted eagle ray) >> comma inserted after 100 
    * 140 (Squirrelfish a) >> comma inserted after 100 
    * 141 (Squirrelfish b) >> comma inserted after 100 
    * 143 (Squirrelfish c) >> comma inserted after 100 
    * 156 (Stomatopods II) >> comma inserted after 100 
    * 164 (Wrasse c) >> comma inserted after 100 
    * 166 (Wrasse d) >> comma inserted after 100 
    * 149 (Wrasse b) >> comma inserted after 100 
    * 180 (Yellowfin mojarra) >> comma inserted after 100 
* Remaining issues:
    * 211 (Grouper a) >> 1.09E+20. What is this value?
    * 204 (Snapper b) >> 1.29E+23. What is this value?
    * What means a value of '0' in column 'prey'? Guild numeriation starts with '1'...

## Load data

Clear R environment and load libraries.
```{r}
# Clear workspace.
rm(list=ls())

# Load libraries.
library(igraph)
library(RCurl)
library(readxl)
```

Load the data from GitHub straight into R. This is more convenient for everyone because one does not have to always change the local file path...

### Serengeti food web
```{r}
# Species in the Serengeti food web.
dd <-  getURL("https://raw.githubusercontent.com/opetchey/RREEBES/master/MORALES-CASTILLA_etal_2015_TREE/Data/Serengeti%20data/journal.pcbi.1002321.s004.CSV",
             cainfo = system.file("CurlSSL", "cacert.pem", package = "RCurl"))
dd.01 <- read.csv(text = dd)
# Feeding links in the Serengeti food web.
dd <-  getURL("https://raw.githubusercontent.com/opetchey/RREEBES/master/MORALES-CASTILLA_etal_2015_TREE/Data/Serengeti%20data/journal.pcbi.1002321.s005.CSV",
             cainfo = system.file("CurlSSL", "cacert.pem", package = "RCurl"))
dd.02 <- read.csv(text = dd)
# Consensus partition.
dd <-  getURL("https://raw.githubusercontent.com/opetchey/RREEBES/master/MORALES-CASTILLA_etal_2015_TREE/Data/Serengeti%20data/journal.pcbi.1002321.s006.CSV",
             cainfo = system.file("CurlSSL", "cacert.pem", package = "RCurl"))
dd.03 <- read.csv(text = dd)
# Link densities between groups in the consensus partition.
dd <-  getURL("https://raw.githubusercontent.com/opetchey/RREEBES/master/MORALES-CASTILLA_etal_2015_TREE/Data/Serengeti%20data/journal.pcbi.1002321.s007.CSV",
             cainfo = system.file("CurlSSL", "cacert.pem", package = "RCurl"))
dd.04 <- read.csv(text = dd)
```

Calculate the number of species and the number of potential links one could expect from this.
```{r}
# Species only in Cuban food web.
n.species.Serengeti       <- length(dd.01[,1])
n.links.species.Serengeti <- n.species.Serengeti^2-n.species.Serengeti

```

Note: There are **175 species** in the Serengeti food web. This results in **25760 potential links** ($n^2-n$).


### Cuban coral reef food web (Morales-Castila et al. only focused on this one).
```{r}
dd <-  getURL("https://raw.githubusercontent.com/opetchey/RREEBES/Morales-Castilla_etal_2015_TREE/MORALES-CASTILLA_etal_2015_TREE/Data/Coral%20reef%20data/857470.item.4%20-%20Guild%20key%2C%20all%20data%20sets.csv",
             cainfo = system.file("CurlSSL", "cacert.pem", package = "RCurl"))
dd.05 <- read.csv(text = dd)
str(dd.05)
# Only data with species from the Cuban food web.
# Guilds only in Cuban food web.
dd.05.a              <- subset(dd.05, Cuba=="x")

# Only the fish pecies in Cuban food web.
n.species.fish.Cuba       <- length(subset(dd.05, Cuba=="x")$Cuba)

# Trophic data for the Cuban food web
dd <-  getURL("https://raw.githubusercontent.com/opetchey/RREEBES/Morales-Castilla_etal_2015_TREE/MORALES-CASTILLA_etal_2015_TREE/Data/Coral%20reef%20data/857470.item.2%20-%20Trophic%20data%20for%20Cuba_sorted.csv",
             cainfo = system.file("CurlSSL", "cacert.pem", package = "RCurl"))
dd.06 <- read.csv(text = dd)
# Guilds in Cuban food web.
n.guilds.Cuba        <- length(dd.06[,1])
n.links.guilds.Cuba  <- n.guilds.Cuba^2-n.guilds.Cuba
```


## How to...?

"*The procedure for building the backbone of an interaction network starts with the identification of species more likely to share similar interactions. The concept is similar to that of modules [28], but we avoid this terminology because modules are usually determined a posteriori and can also refer to simple assemblages of species such as linear food chains or apparent competition [29]. Instead, we define interacting groups based on a priori expectations of interactions. The concept is also analogous to that of guilds [30]. Guilds, however, are restricted to species shar- ing similar resources, and thus do not encompass non- consumptive interactions such as competition or niche construction [27]. A flexible definition of interacting groups based on traits, phylogenies, and geographical distribu- tions would enable combination of heterogeneous information.*" (page 4, bottom-right)

"*Interacting groups of species are defined a priori to simplify the removal of forbidden links. The groups were defined based on the trophic hierarchy of the different species within each eco- system (e.g., primary producers, grazers, small and large carnivores). This process of trophic classification of species led to identification of forbidden links and removal of 􏰁30% of all potential direct links in the coral reef, and 􏰁22% in the Serengeti (e.g., herbivores eating predators; Figure 1).*" (page 5, bottom-left)

"*Refinement of the species groupings was achieved by con- sidering the characteristics of the consumer species (e.g., distinguishing small versus large carnivores in the Seren- geti example, or separating invertebrate feeders, omnivo- rous, and carnivorous fish in the coral reef example; Figure 1).*" (page 5, bottom-right)

## Serengeti foob web analysis

Taking all potential links into account the food webs would look like this:
```{r}
dd.01.a <- expand.grid(dd.01$code, dd.01$code)
# Build the graph.
f <- graph.data.frame(dd.01.a, directed=F, vertices=NULL)
# Now, remove inraspecific trophic interactions - there is no cannibalism. 
dd.01.b <- dd.01.a[which(dd.01.a[,1]!=dd.01.a[,2]),]
```
  
```{r, fig.width=10, fig.height=10}  
# Now, plot the graph.
plot.igraph(f, 
     layout=layout.circle, 
     main="Serengeti food web (no information)",
     vertex.size=8, 
     vertex.color=adjustcolor("#A6B07E", alpha.f=0.6),
     vertex.label.cex=0.5,
     vertex.label.family = "sans",
     edge.width=0.2, 
     edge.color=adjustcolor("#999999", alpha.f=0.3),
     vertex.label.color="#7D1D3F",
     vertex.frame.color="white")
```

This is what the real food web looks like. So far, there is no ordering according to any traits (e.g., nutrition, size...)
```{r}
# Build the graph.
g <- graph.data.frame(dd.02, directed=F, vertices=NULL)
# Display the food web graph.
# tkplot(g)
# plot.igraph(g)

```


```{r, fig.width=10, fig.height=10}
plot.igraph(g, 
#      layout=layout.fruchterman.reingold, 
     layout=layout.circle, 
     main="Serengeti food web (real food web)",
     vertex.size=8, 
     vertex.color=adjustcolor("#A6B07E", alpha.f=0.6),
     vertex.label.cex=0.5,
     vertex.label.family = "sans",
     edge.width=0.2, 
     edge.color=adjustcolor("#333333", alpha.f=0.6),
     vertex.label.color="#7D1D3F",
     vertex.frame.color="white")
```

Once we know about the trophic levels the species/guilds belong to we can use ``layout.reingold.tilford(g, flip.y=FALSE)`` to introduce some hierarchy. For example, if we assign three trophic levels just randomly then the food web can be displayed like this:
```{r, fig.width=10, fig.height=10}
plot.igraph(g, 
#      layout=layout.fruchterman.reingold, 
     layout=layout.reingold.tilford(g, flip.y=FALSE),
     main="Serengeti food web (no trait information)",
     vertex.size=8, 
     vertex.color=adjustcolor("#A6B07E", alpha.f=0.6),
     vertex.label.cex=0.5,
     vertex.label.family = "sans",
     edge.width=2, 
     vertex.label.color="#7D1D3F",
     vertex.frame.color="white")
```
Just to show what is possible with 'igraph'...


## Cuban coral reef food web

### The food web with all its possible links.
```{r}
dd.06.a <- expand.grid(dd.06$Guild.Description, dd.06$Guild.Description)
# Build the graph.
f <- graph.data.frame(dd.06.a, directed=F, vertices=NULL)
# Now, remove inraspecific trophic interactions - there is no cannibalism. 
dd.06.b <- dd.06.a[which(dd.06.a[,1]!=dd.06.a[,2]),]
g <- dd.06.a[which(dd.06.a[,1]==dd.06.a[,2]),]
length(g[,1])
# Should equal the number of potential links, 53.130.
length(g[,1])^2-length(g[,1])
```

```{r, fig.width=10, fig.height=10}
# Now, plot the graph.
plot.igraph(f, 
     layout=layout.circle, 
     main="Cuban coral reef food web (no information)",
     vertex.size=8, 
     vertex.color=adjustcolor("#A6B07E", alpha.f=0.6),
     vertex.label.cex=0.5,
     vertex.label.family = "sans",
     edge.width=0.2, 
     edge.color=adjustcolor("#999999", alpha.f=0.3),
     vertex.label.color="#7D1D3F",
     vertex.frame.color="white")
```

### The food web accounting trophic identity.
This is the point where we stopped. Gian Marco put together the information about who eats whome in the food web, that is, which species really preys upon which other species. This has to be synchronized, yet.


## General conclusion and comments

* The study is quite hard to reproduce for mainly two reasons:
  * The way data is provided is not sufficient for reproduction (to put it nicely). Errors, missing information and poor organization stopped us.
  * We could not figure out how to get from all-links food webs to the more realistic ones given the data that is provided. We had the (possibly wrong) impression that this study is somehow circular in a way that one uses the information about the food web that one wants to reproduce. 








