---
title: "Reproduce analyses of Lague et al.s (2014)"
author: "Alejandra, Marco et al."
date: "26 May 2015"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

# Introduction

This is an attempt to reproduce the anaylses presented in the paper *Parasitism alters three power laws of scaling in a metazoan community: Taylor’s law, density-mass allometry, and variance-mass allometry*, by Lagrue, Poulin, and Cohen ([http://www.pnas.org/content/112/6/1791](http://www.pnas.org/content/112/6/1791)). Data are [here](http://www.pnas.org/content/suppl/2014/12/30/1422475112.DCSupplemental/pnas.1422475112.sd01.txt) as a txt, but need to be saved as csv.

* Here include some text about what is presented below in the basic reproduction, and what additional was done and where that is described.

# First get the raw data into R and tidy it.

To do.

```{r}
# clean R's memory:
rm(list=ls())
# Load packages:
library(tidyr)
library(dplyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(RCurl)
library(reshape2)

# Load data:
dd <- read.csv(text=getURL("https://raw.githubusercontent.com/marcoplebani85/RREEBES/Lagrue/Lagrue_etal_2014_PNAS/data/Lague2014.csv"), header=T)
```

# Tables

## Reproducing Table 1

> This is how to quote in Markdown.

To do.

```{r}
#R code here
```

## Reproducing Table 2

> This is how to quote in Markdown.

To do.

```{r}
#R code here
```

# Figures

## Reproducing Fig. 1

> This is how to quote in Markdown.

To do.

```{r}
#R code here
```

## Reproducing Fig. 2

> This is how to quote in Markdown.

To do.

```{r}
#R code here
```

## Reproducing Fig. 3

> This is how to quote in Markdown.

To do.

```{r}
#R code here
```