---
title: "Reproduce anaylses of Lague et al (2014)"
author: "Alejandra, Marco et al."
date: "26 May 2015"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

# Introduction

This is an attempt to reproduce the anaylses presented in the paper 

[Lagrue, C., Poulin, R. & Cohen, J.E. (2014) . Proceedings of the National Academy of Sciences of the United States of America, 112, 1791–1796.]()
*Parasitism alters three power laws of scaling in a metazoan community: Taylor’s law, density-mass allometry, and variance-mass allometry*, by Lagrue and others ([http://www.pnas.org/content/112/6/1791](http://www.pnas.org/content/112/6/1791)). Data are on [Dryad](http://datadryad.org/resource/doi:10.5061/dryad.sj2hv).

* Here include some text about what is presented below in the basic reproduction, and what additional was done and where that is described.

# First get the raw data into R and tidy it.

To do.

```{r}
# clean R's memory:
rm(list=ls())
# Load packages:
library(tidyr)
library(dplyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(RCurl)
library(reshape2)

# Load data:
spp.abund <- read.csv(text=getURL("https://raw.githubusercontent.com/opetchey/RREEBES/master/Beninca_etal_2008_Nature/data/species_abundances_original.csv"), skip=7, header=T)

# Reproducing Table 1

> This is how to quote in Markdown.

To do.

```{r}
#R code here
```

# Reproducing Table 2

> This is how to quote in Markdown.

To do.

```{r}
#R code here
```

## Reproducing Fig.1

> This is how to quote in Markdown.

To do.

```{r}
#R code here
```

## Reproducing Fig.2

> This is how to quote in Markdown.

To do.

```{r}
#R code here
```

## Reproducing Fig.2

> This is how to quote in Markdown.

To do.

```{r}
#R code here
```