---
title: "Vellend"
author: "Ale, Owen"
date: "4 October 2016"
output: html_document
---

# Introduction


This is an attempt to reproduce the anaylses presented in the paper "Global meta-analysis reveals no net change in local-scale plant biodiversity over time", from Vellend et al. 2013

Biodiversity reduction in small study plots can bring problems to ecosystem function. However, lots of studies have looked at biodiversity loss in experiments while assuming that in nature loss is actually happen at those small scales. The authors tried to test that assumption, by doing a meta-analysis with 168 studies of local-scale non experimental vegetation plots facing disturbance. They showed that mean temporal change in species diversity over periods of 261 years was not different from zero. They found invasive species have a tendency to reduce biodiversity and succession after disturbances show a tendency to increase biodiversity, but other distinctions among studies showed little effect on temporal richness modifications. 

Species involved: all major vegetation types
Median plot area: 44 m2 - specially designed to match spatial scale of BDEF experiments
Effect sizes used: log ratio of species diversity per decade (log(Species richness at the end of study/Species richness at the beginning of study)/time expressed in decades)



# Figures we will intend to reproduce: 

Figure 1: Localize studies in a map based on their coordinates
Figure 2: Number of studies vs effect sizes/decade 
Figure 3: Effect sizes/decade of subsets of studies per Habitat, Continent and Driver. 

Figure S1 and Table S2 in supplementary material: show similar results with subsets of studies that used diversity or eveness as indices instead of species richness. 
Figure S2: Number of studies vs raw effect sizes (not devided by decade)






# Method/Analyses in original paper: 
Criteria for the search of papers is stated in the meta-analysis paper itself and this bibliography search will not be reproduced.
Analyses were done with species richness from 155 studies comprising data from 318 data sets. For each data set a raw effect size was calculated as log ratio of mean species richness in the last and first years of survey. The raw log ratio was standarized to a common decadal time scale, by dividing by the time interval between points. 

The analyses were carried in 2 ways: i) Bayesian analysis of multilevel model, ii) Fixed effect meta-analysis with each study weighted by the square root of the sample size. 
Same analyses were carried for eveness and diversity except for the evaluation of predictor variables, which was omitted due to small sample size. Analyses where done with OepnBUGS in R package, but we will attempt them with the "metafor" package. 

# Method/Analyses in our reproduction: 
Choice of model to use
A) We try first a fixed effects meta-analysis (without weight of the studies). By using fixed effects we assume that there is one true effect size that is shared by all the included studies, so the combined effect estimates a common effect size. Large studies will be given a lot of weight and small studies will be given less weight. The source of error in our estimate is the random error within studies (Borestein et al 2007). 

B) We then try a random effects model, where we allow that the true effect could vary from study to study, so the combined effect estimates the mean effect of the distribution of study effects. The weights assigned under random effects model are more balanced, because each study is estimating a different effect size (from one study to estimate the true effect in a specific populaion and for all the true effects used to estimate the mean of the true effects). Therefore, our estimate depends on the number of subjects within studies and from the total number of studies (Borestein et al 2007)

R code for reproduction

*Figure 1: map points of studies

```{r}
#read data
install.packages("RCurl")
library(RCurl)
setwd("~/Documents/RREEBES/Vellend")
dat <- read.table(text=getURL("https://raw.githubusercontent.com/opetchey/RREEBES/master/Vellend/vellend%20analyses%20reproduction/vellend_original_data.txt"), check.names=FALSE, header=TRUE) 

#make map of the world and lay points from data coordinates
install.packages("rworldmap")
install.packages("rworldxtra")
library(rworldmap)
newmap <- getMap(resolution = "low")
plot(newmap)
points(dat$Longitud, dat$Latitude, col= "red", cex = .6)

#another (nicer) way - world map - legend missing
library("ggmap")
library(maptools)
library(maps)

map("world", fill=TRUE, col="white", bg="white", ylim=c(-60, 90), mar=c(0,0,0,0))
points(dat$Longitud, dat$Latitude, col=dat$Habitat, pch=16, cex=0.8)
title("Figure 1: Coordinates from studies")
legend("bottomleft", horiz = TRUE, col=dat$Habitat) # not working legend


##failed attempt with ggplot2
#  library(ggplot2)
#  world <- map_data("world")
#  worldmap <- ggplot(world, aes(x=long, y=lat, group=group)) 
#  + geom_path() 
#  + scale_y_continuous(breaks=(-2:2) * 30) 
#  + scale_x_continuous(breaks=(-5:5) * 45) 
#  worldmappoints<- worldmap + (geom_point(aes(x = dat$Longitud, y = #  dat$Latitude, colour=dat$Habitat), data=dat)
#  plot(worldmap)



#make map of USA and Europe with points - this package uses Google maps
install.packages("ggmap")
library(ggmap)
mapEu <- get_map(location = 'Europe', zoom = 4)
mapPointsEu <- ggmap(mapEu) + geom_point(aes(x = dat$Longitud, y = dat$Latitude, colour=dat$Habitat), data = dat, alpha = 1)
mapPEu<- mapPointsEu + labs(colour="Habitat")
plot(mapPEu)

mapUSA<-get_map(location="United States of America", zoom=4)
mapPointsUSA <- ggmap(mapUSA) + geom_point(aes(x = dat$Longitud, y = dat$Latitude, colour= dat$Habitat), data = dat, alpha = 1)
mapPUSA<- mapPointsUSA + labs(colour="Habitat")
plot(mapPUSA)

```


*Load and understand data, exclude NA rows in species richness

```{r}
#clear R, load packages, load data
rm(list=ls())
install.packages("metafor")
library("metafor")
install.packages("RCurl")
library(RCurl)

setwd("~/Documents/RREEBES/Vellend")
dat <- read.table(text=getURL("https://raw.githubusercontent.com/opetchey/RREEBES/master/Vellend/vellend%20analyses%20reproduction/vellend_original_data.txt"), check.names=FALSE, header=TRUE) 
View(dat)

#to understand the original data
dat$ratio<- log(dat$SR_Year2_CT/dat$SR_Year1_CT)  #gives you the simple log ratio between means that in original database is named"log_SR_ratio"
dat$ratiostandard<-dat$ratio/dat$Duration  #would give you the effect size standarized that they claim to have used in the paper
View(dat)

#exclude rows with NA in Species Richness column 
dat2<-dat[!is.na(dat$log_SR_ratio),]
View(dat2)

```

Original database has 346 observations (dat). Excluding those with NA in species richness column, I have 318 observations (dat2). 

* Calculate effect size yi and its variance vi
```{r}

#effect size calculation with log transformed ratio of means, to compare with authors calculation in column "log_SR_ratio"
dat2ef <- escalc(measure="ROM", n1i=dat2$N_plots_Year2, n2i=dat2$N_plots_Year1, m1i=dat2$SR_Year2_CT, m2i=dat2$SR_Year1_CT, sd1i=dat2$SR_Year1_STDEV, sd2i=dat2$SR_Year2_STDEV, data=dat2)
View(dat2ef)

```

The authors used "log transformed ratio of means" divided by the standarized time (column: "duration"). Data calculated by us with function escalc using "ROM" fits original data ("yi" matches data in "log_SR_ratio"). "escalc" provides a variance associated with this effect size (vi). Note that there are lots of rows with NA in vi, given that there a lots of standard deviation missing in the original data for 1 of the years of the survey. 

Since authors divided the effect size for the duration, I will divide also the variance (vi) for the duration. This is because I need yi and vi for fitting the fixed effects model (function: "rma"), and vi should have gone through the same standarization as yi. 

```{r}

#standarization of effect sizes for the duration of studies
dat2ef$yistand<-dat2ef$yi / dat2ef$Duration
dat2ef$vistand<-dat2ef$vi / dat2ef$Duration

#barplot figure 2: //needs to be fixed for ranges//
bar<- table(dat2ef$yistand)
barplot(bar, main="Figure 2",
   xlab="Standarized effect size") 

#fit fixed effects model to standarized effect sizes
res<-rma(yi=dat2ef$yistand,vi=dat2ef$vistand,method="FE", data=dat2ef)
print(res,digits=3)
```

The model indicates only 108 studies could be used for this analysis (in original study, authors say 155 studies were included for species richness) - estimate: 0.004***. I try now with the non-standarized effect sizes to see if we get 155 studies included, to see if it makes a difference (maybe values are too small when standarized?). 

```{r}
#fit fixed effects model to standarized effect sizes
res2<-rma(yi,vi,method="FE", data=dat2ef)
print(res2,digits=3)
```

The number of studies included is the same, and the estimate goes up to 0.08***. The low estimate matches conclusions from original articles, although the significance would mean that there is a (really low) significant effect of treatment in species richness. Heterogeneity is highly significant and large in both studies (taking raw or standarized effect sizes).

* group breakdown analyses and meta-regression

```{r}
```

# Results in original paper:
From authors:
The authors found there were increases in biodiversity when original authors had taken succession following major disturbances and to a lesser extent to stop in grazing. A tendency to see declines in species richness were found when authors took invasive species and climate change as drivers, but the 95% credible intervals overlapped zero. 
Temporal change in species richness was independent of plot area, temporal duration and latitude and longitude of a given study. Factors such as spatial constrains and stabilizing species interactions likely buffer against prolonged local-scale diversity change in either direction, given that changes in richness didnt depend on study duration. 

Note: these results dont apply to global large scale biodiversity loss. Also it doesnt undermines conservation efforts to protect ecosystem services. But it assesses the argument that conservation efforts can be justified because of potential ecoszstem function losses. 

Declines of more than 20% were found only in 8% of the datasets. Declines of less than 20% have negligible effects on ecosystem function according to other papers. 

The argument in the discussion is that when sufficiently profound ecosystem changes occur (eg: changing a rainforest to a parking lot), then the impact of biodiversity change on ecosystem function is negligible compared to the overal change in ecosystem composition (soil, nutrients, management, others). So they excluded studies that had gone through overall changes, to reduce the scope of the studie to those that really state a change in ecosystem function based on solely or mainly biodiversity changes. 

Finally they state that although plant species diversity shows no net directional change, there are tremendous variation in direction and magnitude of biodiversity trends and in the identities of species (species composition), which can indeed have an important consequecne for ecosystem function. 


