---
title: "Reproduce anaylses of Beninca et al (2008)"
author: "Owen Petchey"
date: "29 Jan 2015"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

# Introduction

This is an attempt to reproduce the anaylses presented in the paper *Chaos in a long-term experiment with a plankton community*, by Elisa Benincà and others ([the paper on the Nature website](http://www.nature.com/nature/journal/v451/n7180/abs/nature06512.html)). Details of the methods are in [the Supplement to the Nature paper](http://www.nature.com/nature/journal/v451/n7180/extref/nature06512-s1.pdf).

* Here include some text about what is presented below in the basic reproduction, and what additional was done and where that is described.

The data are available as an Excel file supplement to [an Ecology Letters publication](http://onlinelibrary.wiley.com/doi/10.1111/j.1461-0248.2009.01391.x/abstract). The Excel file contains several datasheets. Two are particularly important, as they are the source of the raw data (one contains original species abundances, the one with the nutrient concentrations). Another datasheet in the ELE supplement contains transformed variables.


# First get the raw data into R and tidy it.

```{r}
rm(list=ls())
library(tidyr)
library(dplyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(RCurl)

spp.abund <- read.csv(text=getURL("https://raw.githubusercontent.com/opetchey/RREEBES/master/Beninca_etal_2008_Nature/data/species_abundances_original.csv"), skip=7, header=T)

spp.abund <- select(spp.abund, -X, -X.1)
spp.abund <- spp.abund[-804:-920,]
str(spp.abund)
```

The Protozoa variable contains some numbers with comman as the decimal separator. This creates a question about what dataset was used for the original analyses, as it could not have been this one.

```{r}
spp.abund$Protozoa <- as.numeric(str_replace(spp.abund$Protozoa, ",", "."))
```

Format the dates as dates

```{r}
spp.abund$Date <- dmy(spp.abund$Date)
```

Ooops... R assumes the experiment was done in the 21st century. Shouldn't matter too much.

Data is in wide format, so change it to long:
```{r}
spp.abund <- gather(spp.abund, "variable", "value", 3:12)
str(spp.abund)
```

Bring in the nutrient data:
```{r}
nuts <- read.csv(text=getURL("https://raw.githubusercontent.com/opetchey/RREEBES/master/Beninca_etal_2008_Nature/data/nutrients_original.csv"), skip=7, header=T)
nuts <- select(nuts, -X, -X.1)
nuts <- nuts[-349:-8163,]
nuts$Date <- dmy(nuts$Date)
nuts <- select(nuts, -NO2, -NO3, -NH4)
nuts <- gather(nuts, "variable", "value", 3:4)
str(nuts)
```

Now put the two datasets together
```{r}
all.data <- rbind(spp.abund, nuts)
```

Now select only the date range used in the Nature paper. From the supplment *The analysis in Benincà et al. (Nature 2008) covered all data from 16/06/1991 until 20/10/1997*. (Remembering dates in the R dataframes are 2090s.)
```{r}
all.data <- filter(all.data, Date>dmy("16/06/2091") & Date<dmy("20/10/2097"))
```

# Reproducing figure 1b through 1g

(No attempt to reproduce Figure 1a, as its a food web diagram.)

First quick go:
```{r}
ggplot(all.data, aes(x=Day.number, y=value)) + geom_line() +
  facet_wrap(~variable, scales="free_y")
```

The code from here on needs modifying, as the object names are Owen's old ones, and need changing to those used above.

Now we add a column that gives the variable types, same as in figure 1b through 1g.
First make a lookup table giving species type:
```{r}
tt <- data.frame(variable=unique(all.data$variable),
                 Type=c("Cyclopoids", "Herbivore", "Herbivore", "Herbivore",
                        "Phytoplankton",  "Phytoplankton", "Phytoplankton",
                        "Detritivore", "Detritivore", "Bacteria", "Nutrient", "Nutrient"))
tt
```

And add the Type variable to the new dataset:
```{r}
all.data <- merge(all.data, tt)
```

First lets set the colours as in the original:
```{r}
species.colour.mapping <- c("Cyclopoids"="pink",
                            "Calanoid.copepods"="red",
                            "Rotifers"="blue",
                            "Protozoa"="green",
                            "Nanophytoplankton"="red",
                            "Picophytoplankton"="black",
                            "Filamentous.diatoms"="green",
                            "Ostracods"="lightblue",
                            "Harpacticoids"="purple",
                            "Bacteria"="black",
                            "Total.dissolved.inorganic.nitrogen"="red",
                            "Soluble.reactive.phosphorus"="black")                            
```

Next change the order of the levels in the Type variable, so plots appear in the same order as in the original figure:
```{r}
all.data$Type <- factor(all.data$Type, levels=c("Cyclopoids", "Herbivore", "Phytoplankton", "Nutrient",
                                    "Bacteria", "Detritivore"))
```


Now a version that doesn't try to recreate the "gap" in the y axes of the original figures:
```{r, warning=FALSE, fig.width=10}
g1 <- qplot(as.numeric(Day.number), value, col=variable, data=all.data) +
  facet_wrap(~Type, ncol=2, scales="free_y") +
  geom_point() + geom_line() +
  scale_colour_manual(values = species.colour.mapping)
g1
```
Looks reasonably good.

Now a version that approximates the "gap", by removing data above it:
```{r, warning=FALSE, fig.width=10}
an2 <- filter(all.data, Type=="Cyclopoids" & value<0.6 |
                Type=="Herbivore" & value<13 |
                Type=="Phytoplankton" & value<1400 |
                Type=="Nutrient" & value<50 |
                Type=="Bacteria" & value<10 |
                Type=="Detritivore" & value<0.7) 
g1 <- qplot(as.numeric(Day.number), value, col=variable, data=an2) +
  facet_wrap(~Type, ncol=2, scales="free_y") +
  geom_point() + geom_line() +
  scale_colour_manual(values = species.colour.mapping)
g1
```
Difficult it look like the data go off the top of the graph in ggplot.

Try logarithmic y-axes:
```{r, warning=FALSE, fig.width=10}
g1 <- qplot(as.numeric(Day.number), log10(value+0.00001), col=variable, data=all.data) +
  facet_wrap(~Type, ncol=2, scales="free_y") +
  geom_point() + geom_line() +
  scale_colour_manual(values = species.colour.mapping)
g1
```

Try fourth root, as this is used in the subsequent stats.
```{r, warning=FALSE, fig.width=10}
g1 <- qplot(as.numeric(Day.number), value^0.25, col=variable, data=all.data) +
  facet_wrap(~Type, ncol=2, scales="free_y") +
  geom_point() + geom_line() +
  scale_colour_manual(values = species.colour.mapping)
g1
```

# Data transformation

Now we need to work with transformed data. Details of the transformation, copied from the Supplmentary information:

> 3. Transformation of the time series. We transformed the original time series, shown in Fig. 1b-g of the main text, to obtain stationary time series with equidistant data and homogeneous units of measurement. The transformation steps are illustrated for the bacteria (Fig. S1).

> First, the time series were interpolated using cubic hermite interpolation, to obtain data with equidistant time intervals of 3.35 days (Fig. S1a).

> Next, because the original time series showed many sharp spikes, the time series were rescaled using a fourth-root power transformation (Fig. S1b). The sharp spikes bias "direct method" estimates of the Lyapunov exponent, because nearby pairs of reconstructed state vectors mostly occurred in the troughs between spikes. The average rate of subsequent trajectory divergence from these pairs is therefore an estimate of the local Lyapunov exponent in the troughs, which may be very different from the global Lyapunov exponent. By making spikes and troughs more nearly symmetric, the power transformation resulted in a much more even spread of nearby state vector pairs across the full range of the data for all functional groups in the food web. The transformation is also useful for fitting nonlinear models of the deterministic skeleton (used for nonlinear predictability and indirect method estimates of the Lyapunov exponent), which was done by least squares and therefore is most efficient when error variances are stabilized. Fourth-root transformation is intermediate between the square-root transformation that would approximately stabilize the measurement error variance in count data from random subsamples, and the log transformation that is usually recommended for stabilizing process noise variance due to stochastic variation in birth and death rates.

> The time series were then detrended using a Gaussian kernel with a bandwidth of 300 days (red line in Fig. S1b), to obtain stationary time series. Most species did not show long-term trends, except for the bacteria, detritivores (ostracods and harpacticoid copepods), dissolved inorganic nitrogen and soluble reactive phosphorus. One possible explanation for these trends in the microbial loop could be the slow accumulation of refractory organic material in the mesocosm, but we have not measured this component.

> Finally, the time series were linearly rescaled to have zero mean and a standard deviation of 1 (Fig. S1c).

> The time series of cyclopoid copepods, protozoa, filamentous diatoms, harpacticoid copepods and ostracods contained long sequences of zero values. This does not imply that these species were absent from the food web during these periods, but that their concentrations were below the detection limit. Time series dominated by many zeros can bias the statistical analysis. Therefore, these time series were shortened to remove long sequences of zero values, before the data transformation. The transformed data of all species in the food web are shown in Figure S2.

The ELE supplement contains the raw data and the transformed data, in separate data sheets. I (Owen) also got the interpolated data from Stephen Ellner directly.

And do the spline interpolation
```{r}
mt <- plyr::dlply(all.data,
                  "Species",
                  function(xx) spline(x=xx$Day.number,
                                      y=xx$Abundance,
                                      xout=xout,
                                      method="fmm")$y)
mt <- as.data.frame(mt)
mt <- cbind(Day.number=xout, mt)
mt <- gather(mt, Species, Abundance, 2:13)
```

And compare with the original data
```{r}
soi <- "Bacteria"
g1 <- ggplot(filter(all.data, Species==soi), aes(x=Day.Number, y=Abundance)) +
  facet_wrap(~Species, ncol=2, scales="free_y") +
  geom_line(size=0.5, col="red") +
  scale_colour_manual(values = species.colour.mapping)
g2 <- geom_line(data=filter(mt, Species==soi), aes(x=Day.number, y=Abundance), size=0.25)
g1 + g2
```


# Create dataset with zeros removed


# Spectral analyses

To be done.

# Reproducing Table 1 using ELE supplement data.

To be done.

# Predictability (Figure 2)

To be done.

# Lyapunov exponents by direct method (Figure 3)

To be done.

# Lyapunov exponents by indirect method

To be done.


