---
title: "Reproduce anaylses of Beninca et al (2008)"
author: "Owen Petchey"
date: "29 Jan 2015"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

# Introduction

This is an attempt to reproduce the anaylses presented in the paper *Chaos in a long-term experiment with a plankton community*, by Elisa Benincà and others ([the paper on the Nature website](http://www.nature.com/nature/journal/v451/n7180/abs/nature06512.html)). Details of the methods are in [the Supplement to the Nature paper](http://www.nature.com/nature/journal/v451/n7180/extref/nature06512-s1.pdf).

* Here include some text about what is presented below in the basic reproduction, and what additional was done and where that is described.

The data are available as an Excel file supplement to [an Ecology Letters publication](http://onlinelibrary.wiley.com/doi/10.1111/j.1461-0248.2009.01391.x/abstract). The Excel file contains several datasheets. Two are particularly important, as they are the source of the raw data (one contains original species abundances, the one with the nutrient concentrations). Another datasheet in the ELE supplement contains transformed variables.


# First get the raw data into R and tidy it.

```{r}
rm(list=ls())
library(tidyr)
library(dplyr)
library(lubridate)
library(stringr)
library(ggplot2)

spp.abund <- read.csv("~/Dropbox (Dept of Geography)/RREEBES/Beninca_etal_2008_Nature/data/species_abundances_original.csv", skip=7, header=T)
spp.abund <- select(spp.abund, -X, -X.1)
spp.abund <- spp.abund[-804:-920,]
str(spp.abund)
```

The Protozoa variable contains some numbers with comman as the decimal separator. This creates a question about what dataset was used for the original analyses, as it could not have been this one.

```{r}
spp.abund$Protozoa <- as.numeric(str_replace(spp.abund$Protozoa, ",", ".""))
```






# Reproducing figure 1b through 1g

To be done.

# Data transformation

To be done.

# Spectral analyses

To be done.

# Reproducing Table 1 using ELE supplement data.

To be done.

# Predictability (Figure 2)

To be done.

# Lyapunov exponents by direct method (Figure 3)

To be done.

# Lyapunov exponents by indirect method

To be done.


