---
title: "Reproduction of ecosystem model in *Oxic-anoxic regime shifts mediated by feedbacks between biogeochemical processes and microbial community dynamics* by Bush et al (2017) Nature Communications"
author: "Owen Petchey (and others?)"
date: "9/28/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(deSolve)
```

# Introduction

The aim is to reproduce the model based results / figures.

Here is the original publication:
https://www.nature.com/articles/s41467-017-00912-x

And here is the supplementary information, including the table of parameter values:
https://static-content.springer.com/esm/art%3A10.1038%2Fs41467-017-00912-x/MediaObjects/41467_2017_912_MOESM1_ESM.pdf

Note that below I go with the notation from the table S1 and not the main text. However there is only one small difference between these: In the main text the notation is simplified to y_S_PB (which is y_SR_PB in Table S1) and y_S_SB (which is y_SO_SB in Table S1).


# Reproduce the model

Growth rate function on one or two resources, X, and Y:
```{r}
growth1 <- function(X, g_max, k_X)
  g_max * ( X / (k_X + X) )
growth2 <- function(X, Y, g_max, k_X, k_Y)
  g_max * ( X / (k_X + X) ) * ( Y / (k_Y + Y))
```

Inhibition function:
```{r}
inhibition <- function(X, h_X)
  1 / (1 + (X / h_X))
```

The rate equations:
```{r}
micro_mod <- function(t, state, parameters) {
  
  with(as.list(c(state, parameters)),{
    
    # rates of change
    N_CB_growth_rate <- growth1(P, g_max_CB, k_CB_P) * inhibition(SR, h_SR_CB) * N_CB 
    N_CB_mortality_rate <- m_CB * N_CB
    N_CB_rate <- N_CB_growth_rate - N_CB_mortality_rate
    
    N_PB_growth_rate <- growth2(P, SR, g_max_PB, k_PB_P, k_PB_SR) * inhibition(O, h_O_PB) * N_PB
    N_PB_mortality_rate <- m_PB * N_PB
    N_PB_rate <- N_PB_growth_rate - N_PB_mortality_rate
    
    N_SB_growth_rate <- growth2(P, SO, g_max_SB, k_SB_P, k_SB_SO) * inhibition(O, h_O_SB) * N_SB
    N_SB_mortality_rate <- m_SB * N_SB
    N_SB_rate <- N_SB_growth_rate - N_SB_mortality_rate
    
    SO_rate <- 1 / Y_SR_PB * N_PB_growth_rate -
      1 / Y_SO_SB * N_SB_growth_rate +
      c * O * SR +
      a_S * (back_SO - SO) 
    
    SR_rate <- - 1 / Y_SR_PB * N_PB_growth_rate +
      1 / Y_SO_SB * N_SB_growth_rate -
      c * O * SR +
      a_S * (back_SR - SR) 
    
    O_rate <- P_CB * N_CB_growth_rate -
      c * O * SR +
      a_O * (back_O - O)
    
    P_rate <- - 1 / y_P_CB * N_CB_growth_rate -
      1 / y_P_PB * N_PB_growth_rate -
      1 / y_P_SB * N_PB_growth_rate +
      a_P * (back_P - P)
    
    # return the rate of change
    list(c(N_CB_rate, N_PB_rate, N_SB_rate,
           SO_rate, SR_rate, O_rate, P_rate))
    
  })
  
}
```

Parameter values:
```{r}
parameters <- c(g_max_CB = 0.05,
                g_max_PB = 0.07,
                g_max_SB = 0.1,
                
                k_PB_SR = 10,
                k_SB_SO = 5,
                k_CB_P = 0.2,
                k_PB_P = 0.5,
                k_SB_P = 0.5,
                
                h_SR_CB = 300,
                h_O_PB = 100,
                h_O_SB = 100,
                
                Y_SO_SB = 3.33e7,  
                Y_SR_PB = 1.25e7, ## Y_S_PB in the main text
                y_P_CB = 1.67e8,
                y_P_PB = 1.67e8,
                y_P_SB = 1.67e8,
                
                P_CB = 6e-9,
                
                m_CB = 0.02,
                m_PB = 0.028,
                m_SB = 0.04,
                
                a_S = 0.001,
                a_O = 8e-4,
                a_P = 0.01,
                
                back_SR = 300,
                back_SO = 300,
                back_O = 300,
                back_P = 9.5,
                
                c = 4e-5)

times <- seq(0, 5000, by = 1)

```

# Figure 2a, b

```{r}
state <- c(N_CB = 5e1,
           N_PB = 1e7,
           N_SB = 1e7,
           SO = 300,
           SR = 300,
           O = 1e1,
           P = 1e1)

out <- as.data.frame(ode(y = state,
                         times = times,
                         func = micro_mod,
                         parms = parameters,
                         method = "radau"))
out %>%
  gather(species, quantity, 2:ncol(out)) %>%
  mutate(var_type=ifelse(str_sub(species, 1, 1)=="N", "Organism", "Substrate"),
         trans_quantity=ifelse(var_type=="Organism", log10(quantity), quantity)) %>%
  ggplot(aes(x=time/24, y=trans_quantity, col=species)) +
  facet_wrap(~var_type, scales = "free", ncol = 1) +
  geom_line()
```

Population dynamics seem as in the paper.
**Resource dynamics fluctuate more than in the paper.**

# Figure 2c, d

```{r}
state <- c(N_CB = 1e8,
           N_PB = 1e2,
           N_SB = 1e2,
           SO = 500,
           SR = 50,
           O = 300,
           P = 4)
out <- as.data.frame(ode(y = state,
                         times = times,
                         func = micro_mod,
                         parms = parameters,
                         method = "radau"))
out %>%
  gather(species, quantity, 2:ncol(out)) %>%
  mutate(var_type=ifelse(str_sub(species, 1, 1)=="N", "Organism", "Substrate"),
         trans_quantity=ifelse(var_type=="Organism", log10(quantity), quantity)) %>%
  ggplot(aes(x=time/24, y=trans_quantity, col=species)) +
  facet_wrap(~var_type, scales = "free", ncol = 1) +
  geom_line()

```

Population and resource dynamics see as those in the paper.

# Figure 3

```{r eval=TRUE}
get_final_states <- function(x) {
  #print(x)
  state["N_CB"] <- x["N_CB"]
  parameters["a_O"] <- x["a_O"]
  #print(state)
   as.data.frame(ode(y=state,
                         times = times,
                         func = micro_mod,
                         parms = parameters,
                         method = "radau"))[2,-1]
}


times <- seq(0, 5000, by = 5000)
state <- c(N_CB = NA,
           N_PB = 1e8,
           N_SB = 1e8,
           SO = 250,
           SR = 350,
           O = 150,
           P = 9.5)

expt <- expand.grid(N_CB=10^seq(0, 10, length=20),
                   a_O = 10^seq(-10, -2, length=20))

result <- apply(expt, 1, function(x) get_final_states(x)) %>%
                          tibble() %>%
                          unnest() %>%
                          mutate(initial_N_CB=expt$N_CB,
                                 a_O=expt$a_O)

ggplot(result, aes(x=log10(a_O), y=(N_CB))) +
  geom_point()
ggplot(result, aes(x=log10(a_O), y=(log10(O)))) +
  geom_point()

```

**This reproduction of the model is not able to reproduce the single stable state at low levels of oxygen diffusivity.**
