---
title: "Huisman_Weissing_2001_report"
author: "Frank Pennekamp"
date: "10/28/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve) 
library(plot3D)
```

Here I aim to reproduce the paper "Fundamental unpredictability in multispecies competition" by [Huisman & Weissing (2001)](http://www.jstor.org/stable/10.1086/319929?seq=1#page_scan_tab_contents).

> Abstract: One of the central goals of ecology is to predict the distribution and abundance of organisms. Here, we show that, in ecosystems of high biodiversity, the outcome of multispecies competition can be fundamentally unpredictable. We consider a competition model widely applied in phytoplankton ecology and plant ecology in which multiple species compete for three resources. We show that this competition model may have several alternative outcomes, that the dynamics leading to these alternative outcomes may exhibit transient chaos, and that the basins of attraction of these alternative outcomes may have an intermingled fractal geometry. As a consequence of this fractal geometry, it is impossible to predict the winners of multispecies competition in advance.

The goal of this reproduction are:     
1. learn to set up a competition model based on R theory     
2. reproduce at least figures 1 to 4    

Two models based on the same principles are used in the paper: a 5 species model with competition over three limiting resources, and an 8 species competition model (where three additional species invade the 5 species community).

## Figure 1 
(uses a 8 species community, will be added later)

Let's first define and simulate the 5 species community:

```{r 5 species model, echo=F}

r_star_5 <- function(time, state, pars){ with(as.list(parms), {
  
  rm(list=ls())
  
  # initial conditions for all species and resources
  #N <- rep(0.1,2)
  #S <- rep(10, 3) # micro mol per litre
  
  R <- state[6:8]
  N <- state[1:5]
  S <- rep(10, 3)
  
  # growth rate of all species per day
  r <- rep(1,5)
  
  # mortality (m = D) per day
  m <- rep(0.25,5)
  # resource turnover
  D <- 0.25
  
  # half saturation content for each species on each resource (row = resources, column = species)
  K <- matrix(c(0.2, 0.05, 1, 0.05, 1.2, 0.25, 0.1, 0.05, 1, 0.4, 0.15, 0.95, 0.35, 0.1, 0.05), nrow=3, byrow = T)
  
  # content matrix (row = resources, column = species)
  C <- matrix(c(0.2, 0.1, 0.1, 0.1, 0.1, 0.1, 0.2, 0.1, 0.1, 0.2, 0.1, 0.1, 0.2, 0.2, 0.1), nrow=3, byrow = T)
  
  # species specific growth rates as a function of resource availability
  # maximum growth rate determined by Liebig's law of minimum
  
  r <- rep(1,3)
  r1 <- r[1]*(R[1]/(K[1,1]+R[1]))
  r2 <- r[2]*(R[2]/(K[2,1]+R[2]))
  r3 <- r[3]*(R[3]/(K[3,1]+R[3]))
  mu1 <- min(r1,r2,r3)
  
  r <- rep(1,3)
  r1 <- r[1]*(R[1]/(K[1,2]+R[1]))
  r2 <- r[2]*(R[2]/(K[2,2]+R[2]))
  r3 <- r[3]*(R[3]/(K[3,2]+R[3]))
  mu2 <- min(r1,r2,r3)
  
  r <- rep(1,3)
  r1 <- r[1]*(R[1]/(K[1,3]+R[1]))
  r2 <- r[2]*(R[2]/(K[2,3]+R[2]))
  r3 <- r[3]*(R[3]/(K[3,3]+R[3]))
  mu3 <- min(r1,r2,r3)
  
  r <- rep(1,3)
  r1 <- r[1]*(R[1]/(K[1,4]+R[1]))
  r2 <- r[2]*(R[2]/(K[2,4]+R[2]))
  r3 <- r[3]*(R[3]/(K[3,4]+R[3]))
  mu4 <- min(r1,r2,r3)
  
  r <- rep(1,3)
  r1 <- r[1]*(R[1]/(K[1,5]+R[1]))
  r2 <- r[2]*(R[2]/(K[2,5]+R[2]))
  r3 <- r[3]*(R[3]/(K[3,5]+R[3]))
  mu5 <- min(r1,r2,r3)
  
  # species dynamics
  dn1dt <- N[1] * (mu1 - m[1])
  dn2dt <- N[2] * (mu2 - m[2])
  dn3dt <- N[3] * (mu3 - m[3])
  dn4dt <- N[4] * (mu4 - m[4])
  dn5dt <- N[5] * (mu5 - m[5])
  
  loss_R1 <- sum( 
    (C[1,1] * N[1] * mu1), 
    (C[1,2] * N[2] * mu2), 
    (C[1,3] * N[3] * mu3), 
    (C[1,4] * N[4] * mu4), 
    (C[1,5] * N[5] * mu5)
    )
  
  loss_R2 <- sum( 
    (C[2,1] * N[1] * mu1), 
    (C[2,2] * N[2] * mu2), 
    (C[2,3] * N[3] * mu3), 
    (C[2,4] * N[4] * mu4), 
    (C[2,5] * N[5] * mu5)
  )
  
  loss_R3 <- sum( 
    (C[3,1] * N[1] * mu1), 
    (C[3,2] * N[2] * mu2), 
    (C[3,3] * N[3] * mu3), 
    (C[3,4] * N[4] * mu4), 
    (C[3,5] * N[5] * mu5)
  )
  
  #loss1 <- sum( (C[2,1] * N[1] * mu1), (C[2,2] * N[2] * mu2))
  #loss2 <- sum( (C[2,1] * N[1] * mu1), (C[2,2] * N[2] * mu2))
  #loss3 <- sum( (C[3,1] * N[1] * mu1), (C[2,3] * N[2] * mu2))
  
  # resource dynamics replacement time 
  dr1dt <- D * (S[1] - R[1]) - loss_R1
  dr2dt <- D * (S[2] - R[2]) - loss_R2
  dr3dt <- D * (S[3] - R[3]) - loss_R3
  
  
  
  list(c(dn1dt, dn2dt, dn3dt, dn4dt, dn5dt, dr1dt, dr2dt, dr3dt))
})
}

```

## Figure 2 a)

Then simulate dynamics of 5 species with equal starting conditions (0.1) for all competitors and resources (10).

```{r 5 spec comp a, echo=T}

# define parameters for model simulation
parms <- c(r=1) 

# set up starting conditions
initial_cond1 <- c(0.1, 0.1, 0.1, 0.1, 0.1, 10, 10, 10) 

# simulate dynamics for 2000 time steps (days) using Runge-Kutta algorithm and time step of 0.01 days
out1 <- ode(y = initial_cond1 , times = 1:2000, func = r_star_5, parms = parms) 
```

```{r}
matplot(out1[1:2000 , 2:9], type = "l", xlab = "time", ylab = "Conc",
        main = "5 species R* competition", lwd = 2, col=2:9)
legend("topright", c("sp1", "sp2", "sp3", "sp4","sp5", "R1", "R2", "R3"), col = 2:9, lty = 1)
text(x=1000, y = 40, labels = "Species 1, 4 and 5 wins")
```

## Figure 2 b)

Now simulate dynamics of 5 species with equal starting conditions (0.1) except N_2 (0.2) .

```{r 5 spec comp b, echo=F}
# define parameters for model simulation
parms <- c(r=1) 

# set up starting conditions
initial_cond2 <- c(0.1, 0.1, 0.1, 0.2, 0.1, 10, 10, 10) 
 
# simulate dynamics for 2000 time steps (days) using Runge-Kutta algorithm and time step of 0.01 days
out2 <- ode(y = initial_cond2 , times = 1:2000, func = r_star_5) 
```

```{r}
matplot(out2[1:2000 , 2:9], type = "l", xlab = "time", ylab = "Conc",
        main = "5 species R* competition", lwd = 2, col=2:9)
legend("topright", c("sp1", "sp2", "sp3", "sp4","sp5", "R1", "R2", "R3"), col = 2:9, lty = 1)
text(x=1000, y = 40, labels = "Species 1, 2 and 3 wins")
```

```{r, echo=F}
lines3D(z = out1[1:750,2],
         x = out1[1:750,5],
         y = out1[1:750,3]*-1,  col="darkgreen", main = "Attractor for above time series", plot=F)

lines3D(z = out1[750:2000,2],
         x = out1[750:2000,5],
         y = out1[750:2000,3]*-1,  col="blue",  plot=F, add=T)

lines3D(z = out2[1:500,2],
         x = out2[1:500,5],
         y = out2[1:500,3]*-1,  col="purple", plot=F, add=T)

lines3D(z = out2[500:2000,2],
         x = out2[500:2000,5],
         y = out2[500:2000,3]*-1,  col="blue", add=T, plot=T)

```

Check the outcome of competition based on different starting values:

```{r, eval=F}

## old, not parallel function

# set up initial conditions to vary across species 2 & 4
initial_condition <- expand.grid(sp2=seq(0.00001,2,length.out = 40), sp4=seq(0,2,length.out = 40)) 
initial_condition$sp1 <- initial_condition$sp3 <- initial_condition$sp5 <- 0.1
initial_condition$R1 <- initial_condition$R2 <- initial_condition$R3 <- 10

# re-arrange according to species identity
initial_condition <- initial_condition[, c(5,1,4,2,3,6,7,8) ]

tic <- Sys.time()
basins <- lapply(1:1, function(x) ode(y = as.numeric(initial_condition[x, ]), times = 1:2000, func = r_star_5))
toc <- Sys.time()
toc-tic

save(basins, file = "/Users/frank/Documents/Git projects/RREEBES/Huisman_Weissing_2001/basins2.RData")

# check one output
# matplot(basins[[92]][1:2000 , 2:9], type = "l", xlab = "time", ylab = "Conc",
#         main = "5 species R* competition", lwd = 2, col=2:9)
# legend("topright", c("sp1", "sp2", "sp3", "sp4","sp5", "R1", "R2", "R3"), col = 2:9, lty = 1)

```

Set up initial conditions: 

```{r, eval=F}

# set up initial conditions to vary across species 2 & 4
initial_condition <- expand.grid(sp2=seq(0.00001,2,length.out = resolution), sp4=seq(0,2,length.out = resolution)) 
initial_condition$sp1 <- initial_condition$sp3 <- initial_condition$sp5 <- 0.1
initial_condition$R1 <- initial_condition$R2 <- initial_condition$R3 <- 10

# re-arrange according to species identity
initial_condition <- initial_condition[, c(5,1,4,2,3,6,7,8)]
```

Simulate:

```{r, eval=F}
library(parallel)

# very time consuming code for reproducing figure 3

# Calculate the number of cores
no_cores <- detectCores() - 2

# Initiate cluster
cl <- makeCluster(no_cores, type="FORK")

r_star_5_par <- function(x, time_steps, resolution){

# set up initial conditions to vary across species 2 & 4
initial_condition <- expand.grid(sp2=seq(0.00001,2,length.out = resolution), sp4=seq(0,2,length.out = resolution)) 
initial_condition$sp1 <- initial_condition$sp3 <- initial_condition$sp5 <- 0.1
initial_condition$R1 <- initial_condition$R2 <- initial_condition$R3 <- 10

# re-arrange according to species identity
initial_condition <- initial_condition[, c(5,1,4,2,3,6,7,8) ] 

init <- as.numeric(initial_condition[x, ])

list(ode(y = init, times = 1:time_steps, func = r_star_5))
  
}

clusterExport(cl, list("r_star_5_par", "time_steps", "resolution"))

#basins <- lapply(1:nrow(initial_condition), function(x) ode(y = as.numeric(initial_condition[x, ]), times = 1:time_steps, func = r_star_5))
basins2 <- parLapply(cl, 1:nrow(initial_condition), function(x) {r_star_5_par(x, time_steps, resolution)})
#save(basins2, file = "/Users/frankpennekamp/Github/RREEBES/Huisman_Weissing_2001/basins2.RData")

stopCluster(cl)

```

```{r}
load(file = "/Users/frankpennekamp/Github/RREEBES/Huisman_Weissing_2001/basins2.RData")

# evaluate which species persist
comp_out <- function(x, time, biomass_threshold){ 
  x <- as.data.frame(x)
  log <- x[time,2:6] > biomass_threshold
  winners <- c("sp1", "sp2", "sp3", "sp4", "sp5")[log]
  winners <- stringr::str_c(winners, collapse = ", ")
  paste0( "winners: ", winners)
  
  }
                                       
result <- lapply(basins2, comp_out, time = time_steps, biomass_threshold = 5)
initial_condition$res <- unlist(result)
```


## Figure 3

```{r}
#image(matrix(ifelse(initial_condition$res == "winners: sp1, sp2, sp3", 1,0), byrow = F, ncol = 40),col=c("blue", "yellow"))

library(ggplot2)
ggplot(initial_condition, aes(x=sp2, y=sp4, fill=as.numeric(as.factor(initial_condition$res)))) + geom_tile() + scale_fill_gradient(low = "yellow", high = "blue") + guides(fill=F) + coord_fixed()


```

